Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-17 MS_thesis.tex
Update to reflect changes in ind files
Add bibliography to TOC
Update thesis abstract

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where mechanistic rules are specified for each individual, are more realistic but are rarely studied analytically.\\
\\
In this thesis, I show how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system as a stochastic differential equation (SDE). I show that well-known ‘fundamental theorems’ of evolution, such as the Price equation, are recovered in the infinite population limit. My stochastic formulation reveals a new directional evolutionary force in finite populations that lends itself to a simple biological interpretation and has no analogue in infinitely large populations. As an application, I show how this formalism can be used to study cooperation in finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation, which can be thought of as an infinite-dimensional SDE. In the infinite population limit, I recover well-known frameworks such as Kimura's continuum-of-alleles and adaptive dynamics. My work thus highlights the connections between various equations in ecology and evolution by revealing their common stochastic underpinnings.
}

@@

\part{Discussion \& Outlook}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/06_unification_discussion}
\chapter{Towards a stochastic evolutionary theory}\label{chap_stochasticity}
\input{ind_files/07_stochastic_discussion}
% \chapter{}
% \input{ind_files/chapter_05}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_06}

@@

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}

\printbibliography[title=References]
\end{document}

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Some central questions of population biology relate to the number of distinct variants of an entity that can emerge and coexist in a closed population. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. In this thesis, I use ideas from statistical physics to analytically describe evolving populations from biological first principles. Starting from a density-dependent ‘birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations for how trait frequencies change over time. Well-known relations such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. I then extend this theory to one-dimensional quantitative traits through a ‘stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and adaptive dynamics as special cases. I make no assumptions other than density-dependence and the impossibility of infinite growth, and thus, my derivation is almost universally valid. The resultant equations reveal a new directional evolutionary force, ‘noise-induced selection’, that is particular to finite populations and has been overlooked in standard mathematical formulations of evolution. Noise-induced selection can reverse the direction of evolution predicted by infinite-population models, with implications for simulation studies and real world populations. My work thus derives a set of fundamental equations showcasing the common stochastic underpinnings of several formal structures in population biology and predicts a new evolutionary force unique to finite populations.
}

@@

\part{Discussion \& Outlook}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/06_unification_discussion}
%\chapter{Towards a stochastic evolutionary theory}\label{chap_stochasticity}
%\input{ind_files/07_stochastic_discussion}
% \chapter{}
% \input{ind_files/chapter_08}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_09}

@@

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}

\cleardoublepage
\phantomsection
\printbibliography
\addcontentsline{toc}{part}{References}
\end{document}
